---
title: "2. Plotting fits"
author: "Giulio Caravagna"
date: "`r format(Sys.time(), '%d %B, %Y')`"
institute: "Institute for Cancer Research"
email: "giulio.caravagna@icr.ac.uk"
output: rmarkdown::github_document
always_allow_html: true
vignette: >
  %\VignetteIndexEntry{2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r, message=FALSE, warning=F}
library(mobster)
library(tidyr)
library(dplyr)
```

This vignette describes the plotting functions available in `mobster`. As an example, we use
one of the available datasets where we annotate some random mutations as drivers.

```{r, fig.width=5, fig.height=4}
# Example data where we have 3 events as drivers
example_data = Clusters(mobster::fit_example$best) %>% dplyr::select(VAF)

# Drivers annotation (we selected this entries to have nice plots)
drivers_rows = c(2239, 3246, 3800)

example_data$is_driver = FALSE
example_data$driver_label = NA

example_data$is_driver[drivers_rows] = TRUE
example_data$driver_label[drivers_rows] = c("DR1", "DR2", "DR3")

# Fit and print the data
fit = mobster_fit(example_data, auto_setup = 'FAST')

best_fit = fit$best
print(best_fit)
```

# Model plots

The plot reports some fit statistics, and shows the annotated drivers if any.

```{r, fig.width=5, fig.height=4}
plot(best_fit)
```

One can hide the drivers setting `is_driver` to `FALSE`. 

```{r, fig.width=5, fig.height=4}
copy_best_fit = best_fit 
copy_best_fit$data$is_driver = FALSE 

plot(copy_best_fit)
```

It is possible to annotate further labels to this plot, providing just a VAF value and a `driver_label` value. Each annotation points to the VAF value on the x-axis, and on the corresponding mixture density value for the y-axis.

```{r, fig.width=5, fig.height=4, warning = FALSE}
plot(best_fit,
     annotation_extras = 
       data.frame(
         VAF = .35, 
         driver_label = "Something",
         stringsAsFactors = FALSE)
     )
```

Other visualisations can be obtained as follows
```{r, fig.width=10, fig.height=8, warning = FALSE}
ggpubr::ggarrange(
  plot(best_fit, tail_color = "darkorange"),                  # Tail color
  plot(best_fit, beta_colors = c("steelblue", "darkorange")), # Beta colors
  plot(best_fit, cutoff_assignment = .50),                    # Hide mutations based on latent variables 
  plot(best_fit, secondary_axis = "SSE"),               # Add a mirrored y-axis with the % of SSE
  ncol = 2,
  nrow = 2
)
```

# Fit statistics

You can plot the latent variables, which are used to determine *hard clustering* assignments of the input mutations. A `cutoff_assignment` determines a cut to prioritize assignments above the hard clustering assignments probability. Non-assignable mutations (`NA` values) are on top of the heatmap.

```{r, fig.width=10, fig.height=5}
ggpubr::ggarrange(
  mobster::plot_latent_variables(best_fit, cutoff_assignment = 0),
  mobster::plot_latent_variables(best_fit, cutoff_assignment = 0.4),
  mobster::plot_latent_variables(best_fit, cutoff_assignment = 0.8),
  mobster::plot_latent_variables(best_fit, cutoff_assignment = 0.97),
  ncol = 4,
  nrow = 1
)
```

You can plot a barplot of the mixing proportions, using the same colour scheme for the fit (here, default). The barplot annotates a dashed line by default at 2%.

```{r, fig.width=3, fig.height=3}
plot_mixing_proportions(best_fit)
```

The negative log-likelihood `NLL` of the fit can be plot against the iteration steps, so to check that the trend is decreasing over time.
```{r, fig.width=3, fig.height=3}
plot_NLL(best_fit)
```

A contributor to the `NLL` is the entropy of the mixture, which can be visualized along with the reduced entropy, which is used by `reICL`.
```{r, fig.width=3, fig.height=3}
plot_entropy(best_fit)
```

# Inspecting alternative fits

Alternative models are returned by function `mobster_fit`, and can be easly visualized; the table reporting the scores is a good place to start investigating alternative fits to the data.

```{r, fig.width=15, fig.height=4, message=FALSE, warning=FALSE}
print(fit$fits.table)
```

Top three fits, plot in-line.
```{r, fig.width=15, fig.height=4, message=FALSE, warning=FALSE}
ggpubr::ggarrange(
  plot(fit$runs[[1]]),
  plot(fit$runs[[2]]),
  plot(fit$runs[[3]]),
  ncol = 3, 
  nrow = 1
)
```

The sum of squared error (SSE) between the fit density and the data (binned with bins of size `0.01`), can be plot to compare multiple fits. This measure is a sort of "goodness of fit" statistics.

```{r, fig.width=6, fig.height=3}
# Goodness of fit (SSE), for the top 3 fits.
plot_gofit(fit, TOP = 3)
```

The scores for model selection can also be compared graphically. 

```{r, fig.width=4, fig.height=4}
plot_fit_scores(fit)
```

In the above plot, all the computed scoring functions (`BIC`, `AIC`, `ICL` and `reICL`) are shown. This plot can be used to quickly grasp the best model with, for instance, the default score (`reICL`) is also the best for other scores. In this graphics the red dot represents the best model according to each possible score.

# Model-selection report

A general model-selection report assembles most of the above graphics.

```{r, fig.width=9, fig.height=9, message=FALSE, warning=FALSE}
plot_model_selection(fit, TOP = 5)
```

# Fit animation

It seems useless, but if you want you can animate `mobster` fits using [plotly](https://plotly.com/r/).

```{r, message=F, warning=F, eval=TRUE}
example_data$is_driver = FALSE

# Get a custom model using trace = TRUE
animation_model = mobster_fit(
  example_data, 
  parallel =  FALSE, 
  samples = 3,
  init = 'random',
  trace = TRUE,
  K = 2,
  tail = TRUE)$best

# Prepare trace, and retain every 5% of the fitting steps
trace = split(animation_model$trace, f = animation_model$trace$step)
steps = seq(1, length(trace), round(0.05 * length(trace)))
  
# Compute a density per step, using the template_density internal function
trace_points = lapply(steps, function(w, x) {
  new.x = x
  new.x$Clusters = trace[[w]]
  
  # Hidden function (:::)
  points = mobster:::template_density(
    new.x,
    x.axis = seq(0, 1, 0.01),
    binwidth = 0.01,
    reduce = TRUE
  )
  points$step = w
  
  points
},
x = animation_model)
  
trace_points = Reduce(rbind, trace_points)
  
# Use plotly to create a ShinyApp
require(plotly)
    
trace_points %>%
  plot_ly(
    x = ~ x,
    y = ~ y,
    frame = ~ step,
    color = ~ cluster,
    type = 'scatter',
    mode = 'markers',
    showlegend = TRUE
  )
```
