---
title: "4. Population Genetics statistics"
author: "Giulio Caravagna"
date: "`r format(Sys.time(), '%d %B, %Y')`"
institute: "Institute for Cancer Research"
email: "giulio.caravagna@icr.ac.uk"
output: rmarkdown::github_document
vignette: >
  %\VignetteIndexEntry{4}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r, message=FALSE, warning=F}
library(mobster)
library(tidyr)
library(dplyr)
```

Population Genetics statistics can be extracted from a MOBSTER model. 

```{r}
data('fit_example', package = 'mobster')
print(fit_example$best)

evolutionary_parameters(fit_example)
```

The mutation rate `mu` (cell division units) scaled by the probability of lineage survival $\beta$, $\mu/\beta$, is given by:
\[
\mu/\beta = \dfrac{M} {(\frac{1}{f_\text{min}} - \frac{1}{f_\text{max}})}
\]
Where $f_\text{min}$ is the minimum VAF and $f_\text{max}$ is the maximum, and
$M$ is the number of mutations between $f_\text{min}$ and $f_\text{max}$.

Selection is defined as the relative growth rates of host tumour cell populations ($\lambda h$) vs subclone ($\lambda s$):
\[
1+s= \dfrac{\lambda h}{ \lambda s}
\]

The mathematical details of these computations are described in the main paper, and baesd on the population genetics model of tumour evolutionin Williams et al. 2016 and 2018 (Nature Genetics).

