---
title: "a8_MOBSTERm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{a8_MOBSTERm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



```{r setup}
devtools::install_github("caravagnalab/mobster@mobsterm")
library(mobster)
library(tidyr)
library(dplyr)
```

`MOBSTERm` is a tool that implements a Bayesian model to perform multivariate subclonal deconvolution, allowing to detect neutral, private and selected clonal mutations in multi-region and longitudinal cancer datasets.

# Input data for `MOBSTERm`

The input data consists of a dataframe (or tibble) `x` containing information about somatic mutations (SNVs). Each row represents a mutation, with the following columns:

- **`mutation_id`**: Unique identifier for the mutation.
- **`sample_id`**: Unique identifier for the sample.
- **`NV`**: Number of reads supporting the variant.
- **`DP`**: Depth of sequencing at the mutation site.
- **`purity`**: Purity of the sample.
- **`karyotype`**: Copy number (cn) state of the sample. The karyotype value must be unique for each sample. The cn state must be in the format `major_cn:minor_cn`.

An example of an input dataframe is the following:

...

# Fitting the model
Function `mobsterm_fit()` fits a `MOBSTERm` model.

#### Key parameters
This function has several parameters to customize the fitting procedure.

- **`K_list`**: A vector with the number of mixture components to use. All values of the liist must be positive and strictly greater than 0.
- **`max_iter`**: Maximum number of steps for a variational inference fit. If parameters and loss convergence is not achieved before these steps, the fit is interrupted.
- **`seed`**: A vector with the number of seeds to test for the fit.
- **`lr`**: Learning rate used by the optimizer.
- **`par_threshold`**: Tolerance for parameter convergence. As ELBO oscillations are common in gradient based VI, we will monitor the convergence of all the parameters in the model, the inference stops when (abs(new-old) / abs(old)) < par_threshold for 200 consecutive iterations, for all the parameters.
- **`loss_threshold`**: Tolerance for loss convergence. As ELBO oscillations are common in gradient based VI, we will monitor the convergence of the loss in the model, the inference stops when (abs(new_loss-old_loss) / abs(old_loss)) < loss_threshold for 200 consecutive iterations.

#### Output
The output is in the format..


# Model plots

Here we show how to plot the results of the fit.
In particular we plot the multivariate scatter, the histograms of the marginals, the delta parameter.

# Fit statistics

Here we show how to plot responsibilities, mixing proportions, model selection scores, ELBO, parameters gradients, parameters and loss convergence.

# Inspecting alternative fits

Here we show how to plot alternative model fits.
